### 2015-07-11 rscripts/data_review.Rmd
review refinement

```{r}
max_hours <- 24
min_logins <- 2
```

@@

```{r, echo=FALSE}
ggplot(censor.dt[,list(t_min=min(login),t_max=max(logout)), by=location_id]) +
  theme_bw() +
  aes(x=location_id, ymin=t_min, ymax=t_max) +
  geom_linerange() +
  coord_flip() + labs(y="seconds since start", x="hotspot ordered by start time")
```

@@

```{r, echo=FALSE}
weekly_ave <- censor.dt[,(length(unique(location_id))-2)/(max(login) - min(login))*60*60*24*7]
weekly_count <- censor.dt[,list(creation_week=(min(login_day) %/% 7)), by=location_id][,list(count=.N),by=creation_week]
ggplot(weekly_count) + theme_bw() + aes(x=creation_week, y=count) +
  geom_bar(stat="identity", width=1) +
  geom_hline(y = weekly_ave, color="red")
```

@@

```{r, echo=FALSE}
creations <- setkey(censor.dt[,list(creation_day=min(login_day)), by=location_id][,list(count=.N), by=creation_day][,list(creation_day = creation_day - min(creation_day), count)], "creation_day")
zeros <- data.table(creation_day = 0:max(creations$creation_day), count=0, key = "creation_day")
creations <- merge(creations, zeros, all=TRUE)[,list(count=max(count.x, count.y, na.rm=T)),keyby=creation_day]
ggplot(creations) + theme_bw() + aes(x=creation_day, y=count) +
  stat_smooth(method="lm") + geom_point()
```

@@

Unlike creation, elimination of hotspot IDs seems to be much more variable.  So they are steadily created, but the lifetime after creation is variable.

```{r,echo=FALSE}
# dt <- censor.dt[,list(duration=(max(logout) - min(login))/60/60/24/7), by=location_id]
# h <- hist(dt$duration, 0:ceiling(max(dt$duration)), plot=F)
# ggplot() +
#   theme_bw() +
#   aes(x=duration/60/60/24/7) +
#   geom_bar(binwidth=1) + labs(x="weeks of hotspot life", y="hotspots with that duration")
```

The usage profile of hostspots also varies.  We can use this to categorize "types" of locations using a *k-means* approach.  First let's look at grouping by typical *login* times:

```{r, echo=FALSE}
shft <- 4
orig <- 0:23
hr_shf <- (orig - shft) %% 24
hr_lab <- (orig + shft) %% 24
base <- censor.dt[,list(login_hour=floor(login_time/60/60)), by=location_id][,list(count=.N),keyby=list(location_id, login_hour)][,list(login_hour=hr_shf[login_hour+1],count=count/max(count)),keyby=location_id]
# by max vs by sum
byid <- acast(base, location_id ~ login_hour, value.var = "count", fill = 0)
kmplotter <- function(n) {
  km <- kmeans(byid, n, nstart = 5)
  ggplot(base[,list(location_id,login_hour,count,cluster=factor(km$cluster[location_id]))]) + theme_bw() +
    aes(x = factor(location_id), y = login_hour, fill = count) +
    facet_grid(. ~ cluster, scales = "free") + scale_x_discrete("location", labels="") +
    geom_raster() + scale_y_discrete(name="login hour", labels=hr_lab, breaks=0:23) +
    labs(fill="proportion max")
}
print(kmplotter(4))
print(kmplotter(5))
print(kmplotter(6))
print(kmplotter(7))
print(kmplotter(8))
print(kmplotter(9))
print(kmplotter(10))
```

We can also look at logout times:

```{r, echo=FALSE}
shft <- 4
orig <- 0:23
hr_shf <- (orig - shft) %% 24
hr_lab <- (orig + shft) %% 24
base <- censor.dt[,list(logout_hour=floor(logout_time/60/60)), by=location_id][,list(count=.N),keyby=list(location_id, logout_hour)][,list(logout_hour=hr_shf[logout_hour+1],count=count/max(count)),keyby=location_id]
# by max vs by sum
byid <- acast(base, location_id ~ logout_hour, value.var = "count", fill = 0)
kmplotter <- function(n) {
  km <- kmeans(byid, n, nstart = 5)
  ggplot(base[,list(location_id,logout_hour,count,cluster=factor(km$cluster[location_id]))]) + theme_bw() +
    aes(x = factor(location_id), y = logout_hour, fill = count) +
    facet_grid(. ~ cluster, scales = "free") + scale_x_discrete("location", labels="") +
    geom_raster() + scale_y_discrete(name="logout hour", labels=hr_lab, breaks=0:23) +
    labs(fill="proportion max")
}
print(kmplotter(4))
print(kmplotter(5))
print(kmplotter(6))
print(kmplotter(7))
print(kmplotter(8))
print(kmplotter(9))
print(kmplotter(10))
```

>>>>

```{r, echo=FALSE}
max_hours <- 24
min_logins <- 2
```

@@

```{r, echo=FALSE}
minmax.dt <- censor.dt[,list(t_min=min(login),t_max=max(logout)), keyby=location_id]
ggplot(minmax.dt) +
  theme_bw() +
  aes(x=location_id, ymin=t_min/60/60/24/7, ymax=t_max/60/60/24/7) +
  geom_linerange() +
  coord_flip() + labs(y="weeks since start", x="hotspot")
```

@@

```{r, echo=FALSE}
weekly_ave <- minmax.dt[,
  (length(unique(location_id))-2)/(max(t_max) - min(t_min))*60*60*24*7
]
weekly.dt <- minmax.dt[,
  list(
    creation = (t_min/60/60/24)%/%7,
    shutdown = (t_max/60/60/24)%/%7
  ),
  by = location_id
]
count.dt <- weekly.dt[,
  .N, by = creation
]
ggplot(count.dt) + theme_bw() + 
  aes(x=creation, y=N) +
  labs(x="week", y="hotspots created", color="ave") +
  geom_bar(stat="identity", width=1) +
  geom_hline(y = weekly_ave, color="red")
```

@@

```{r, echo=FALSE}
creations <- setkey(censor.dt[,
  list(creation_day=min(login_day)),
  by=location_id
][,
  .N,
  by=creation_day
][,
  list(creation_day = creation_day - min(creation_day), N)
], creation_day)
zeros <- data.table(
  creation_day = 0:max(creations$creation_day),
  N = 0,
  key = "creation_day"
)
creations <- merge(creations, zeros, all=TRUE)[,
  list(N = max(N.x, N.y, na.rm=T)),
  keyby = creation_day
]
ggplot(creations) + theme_bw() +
  aes(x=creation_day, y=N) +
  stat_smooth(method="lm") + geom_point() +
  labs(x="creation day", y="hotspots created")
```

@@

Unlike creation, elimination of hotspot IDs seems to be much more variable.  While they are steadily created, their lifetime after creation has a long tailed distribution.  This is further complicated by the termination of the data, left censoring many entries.

TODO: break out entries that start near the end of the dataset and highlight their contribution to the spike in low weeks.

```{r,echo=FALSE}
dt <- minmax.dt[,
  list(duration = (t_max-t_min)/60/60/24/7),
  by=location_id
]
# h <- hist(dt$duration, 0:ceiling(max(dt$duration)), plot=F)
ggplot(dt) +
  theme_bw() +
  aes(x=duration) +
  geom_bar(binwidth=1) + labs(x="weeks of hotspot life", y="hotspots with that duration")
```

The usage profile of hostspots also varies.  We can use this to categorize "types" of locations using a *k-means* approach.  First, let's look at grouping by typical *login* and *logout* times:

```{r, echo=FALSE}
shft <- 4
orig <- 0:23
hr_shf <- (orig - shft) %% 24
hr_lab <- (orig + shft) %% 24
base <- censor.dt[,
  list(
    login_hour  = floor(login_time/60/60),
    logout_hour = floor(logout_time/60/60)
  ),
  by = location_id
]
both <- rbind(base[,
  list(login_count=.N),
  keyby=list(location_id, login_hour)
][,
  list(
    hour=hr_shf[login_hour+1],
    count=login_count/max(login_count),
    event=factor("login", levels=c("logout","login"))
  ),
  keyby=location_id
],
base[,
  list(logout_count=.N),
  keyby=list(location_id, logout_hour)
][,
  list(
    hour=hr_shf[logout_hour+1],
    count=logout_count/max(logout_count),
    event=factor("logout", levels=c("logout","login"))
  ),
  keyby=location_id
])
setkey(both, location_id, event, hour)
# by max vs by sum?
byid <- acast(both, location_id ~ hour + event, value.var = "count", fill = 0)

kmplotter <- function(n, castdata, src,
  facet = facet_grid(event ~ cluster, scales = "free"),
  refevent = "login"
) {
    km <- kmeans(castdata, n, nstart = 5)
    cluster.dt <- data.table(
      location_id = 1:src[,max(location_id)],
      cluster = km$cluster, 
      key="location_id"
    )
    peakhours <- src[cluster.dt][,
      list(peak=sum(hour*count)/sum(count)),
      by=list(event, cluster)
    ]
    setkey(peakhours, event, peak)
    ## get the order from login
    ## relabel
    reorder <- peakhours[event == refevent, order(cluster)]
    peakhours[, cluster := reorder[cluster]]
    cluster.dt[,cluster := reorder[cluster]]
  # reorder cluster ids
    ggplot(src[cluster.dt]) +
      theme_bw() +
      aes(x = factor(location_id), y = hour, fill = count) + facet +
      geom_raster() + geom_hline(data=peakhours, mapping=aes(yintercept = peak), color="red") +
      scale_x_discrete("location",  labels="") +
      scale_y_discrete(name="hour", labels=hr_lab, limits=0:23, expand = c(0,0)) +
      labs(fill="prop. max count")
}
# print(kmplotter(3))
# print(kmplotter(4))
print(kmplotter(5, byid, both)) 
# print(kmplotter(6))
# print(kmplotter(7))
# print(kmplotter(8))
# print(kmplotter(9))
# print(kmplotter(10))
# TODO: sort out clustering - either different clustering algorithm, or principled selection of k?
```

An alternative view would be to consider something like total users present in each hour.

TODO: could weight a user contribution to an hour by how much time they are present during it, but seems reasonable to just do counts at the moment.

```{r, echo=FALSE}
if (file.access("extend.RData") != 0) {
  proto <- array(0L, dim = c(1,24), dimnames = list(drop=c(),hours=0:23))
  extend <- cbind(censor.dt, proto)
  extend[, login_hour := floor(login_time/60/60) ][, logout_hour:=floor(logout_time/60/60)]
  extend[, "0"  := ifelse((0  <= logout_hour) & (login_hour <= 0),  1L, 0L)]
  extend[, "1"  := ifelse((1  <= logout_hour) & (login_hour <= 1),  1L, 0L)]
  extend[, "2"  := ifelse((2  <= logout_hour) & (login_hour <= 2),  1L, 0L)]
  extend[, "3"  := ifelse((3  <= logout_hour) & (login_hour <= 3),  1L, 0L)]
  extend[, "4"  := ifelse((4  <= logout_hour) & (login_hour <= 4),  1L, 0L)]
  extend[, "5"  := ifelse((5  <= logout_hour) & (login_hour <= 5),  1L, 0L)]
  extend[, "6"  := ifelse((6  <= logout_hour) & (login_hour <= 6),  1L, 0L)]
  extend[, "7"  := ifelse((7  <= logout_hour) & (login_hour <= 7),  1L, 0L)]
  extend[, "8"  := ifelse((8  <= logout_hour) & (login_hour <= 8),  1L, 0L)]
  extend[, "9"  := ifelse((9  <= logout_hour) & (login_hour <= 9),  1L, 0L)]
  extend[, "10" := ifelse((10 <= logout_hour) & (login_hour <= 10), 1L, 0L)]
  extend[, "11" := ifelse((11 <= logout_hour) & (login_hour <= 11), 1L, 0L)]
  extend[, "12" := ifelse((12 <= logout_hour) & (login_hour <= 12), 1L, 0L)]
  extend[, "13" := ifelse((13 <= logout_hour) & (login_hour <= 13), 1L, 0L)]
  extend[, "14" := ifelse((14 <= logout_hour) & (login_hour <= 14), 1L, 0L)]
  extend[, "15" := ifelse((15 <= logout_hour) & (login_hour <= 15), 1L, 0L)]
  extend[, "16" := ifelse((16 <= logout_hour) & (login_hour <= 16), 1L, 0L)]
  extend[, "17" := ifelse((17 <= logout_hour) & (login_hour <= 17), 1L, 0L)]
  extend[, "18" := ifelse((18 <= logout_hour) & (login_hour <= 18), 1L, 0L)]
  extend[, "19" := ifelse((19 <= logout_hour) & (login_hour <= 19), 1L, 0L)]
  extend[, "20" := ifelse((20 <= logout_hour) & (login_hour <= 20), 1L, 0L)]
  extend[, "21" := ifelse((21 <= logout_hour) & (login_hour <= 21), 1L, 0L)]
  extend[, "22" := ifelse((22 <= logout_hour) & (login_hour <= 22), 1L, 0L)]
  extend[, "23" := ifelse((23 <= logout_hour) & (login_hour <= 23), 1L, 0L)]
} else {
  load("extend.RData")
}
if (file.access("comp.RData") != 0) {
  comp <- extend[,list(
    `0` =sum(`0`),
    `1` =sum(`1`),
    `2` =sum(`2`),
    `3` =sum(`3`),
    `4` =sum(`4`),
    `5` =sum(`5`),
    `6` =sum(`6`),
    `7` =sum(`7`),
    `8` =sum(`8`),
    `9` =sum(`9`),
    `10`=sum(`10`),
    `11`=sum(`11`),
    `12`=sum(`12`),
    `13`=sum(`13`),
    `14`=sum(`14`),
    `15`=sum(`15`),
    `16`=sum(`16`),
    `17`=sum(`17`),
    `18`=sum(`18`),
    `19`=sum(`19`),
    `20`=sum(`20`),
    `21`=sum(`21`),
    `22`=sum(`22`),
    `23`=sum(`23`)
  ),by=location_id]
  save(comp, file="comp.RData")
} else {
  load("comp.RData")
}

comp.short <- melt(comp, id.var="location_id", variable.name = "hour", value.name="count")[count != 0]
comp.short$count <- as.double(comp.short$count)
comp.short$hour <- as.integer(comp.short$hour)
setkey(comp.short, location_id, hour)
comp.short[, count := count/max(count), by=location_id][,event := "usage" ]
compbyID <- acast(comp.short, location_id ~ hour, value.var = "count", fill = 0)

kmplotter(5, compbyID, comp.short, facet = facet_grid(. ~ cluster, scale="free"), refevent="usage")
```

These analyses identify similar kinds of locations based on their aggegrated usage profile.  There is not an obviously preferable approach to the signature we used for clustering -- *i.e.*, login and logout densities *vs.* concurrent usage density -- so we choose the usage-based clusters.

TODO: there probably could be, if we did some statistics to determine best *k*, and then did some information criterion between the two models?  Additionally, compare overlap in clusters?

If we revisit our view of the presence of locations over time in the dataset, what do we see?

```{r,echo=FALSE}
usagekm <- kmeans(compbyID, 8, nstart = 5)$cluster
km.dt <- data.table(location_id=1:length(usagekm), cluster=factor(usagekm), key=c("location_id"))
ggplot(minmax.dt[km.dt]) +
  theme_bw() +
  aes(x=location_id, ymin=t_min, ymax=t_max, color=cluster) +
  geom_linerange() +
  coord_flip() + labs(y="seconds since start", x="hotspot ordered by start time") +
  scale_color_discrete()
```

There are perhaps some trends like-location creation, indicated by the solid color bands.  However, this does not clarify what sort of locations are available throughout the whole duration of the data set.  What if we look at the availability of a type cluster at anytime?

```{r, echo=FALSE}
week.melt <- setkey(melt(weekly.dt, id.var = "location_id", variable.name = "event", value.name = "week"), location_id)[km.dt]
week.melt[, inc := ifelse(event == "creation", 1, -1)]
week.melt <- week.melt[,list(inc = sum(inc)), keyby=list(cluster, week) ]
# ggplot(week.totals[cluster==1]) + aes(x = week, y = agg, fill=factor(cluster), group=cluster) + geom_area(stat="hline")
week.melt.fill <- setkey(Reduce(function(left, right) {
  ul <- left[!(week %in% right[,unique(week)]), unique(week)]
  ur <- right[!(week %in% left[,unique(week)]), unique(week)]
  ulc <- left[,unique(cluster)]
  rbind(left, right, 
    data.table(week=ul, inc=0, cluster=right[,unique(cluster)]),
    data.table(week=rep(ur, times=length(ulc)), inc=0, cluster=rep(ulc, each=length(ur)))
  )
}, Map(function(clus) week.melt[cluster == clus], week.melt[,unique(cluster)])), cluster, week)
week.totals <- week.melt.fill[,list(agg = cumsum(inc), week), keyby=cluster]
clusorder <- week.totals[,list(mx=max(agg)),keyby=cluster][,cluster,keyby=mx]$cluster
week.totals$cluster <- factor(week.totals$cluster, levels=clusorder)
ggplot(week.totals) + aes(x = week, y = agg, fill=cluster, order=cluster) + geom_area(position="fill")
```

This suggests that, at least to a reasonable approximation, the prevalance of cluster types is constant through the range we are considering for the population synthesis.

Now, how do real users use these locations?
